Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-05 paper/main.tex
spell check

  \caption{The left automaton accepts the words % $w\in\{a,b\}^\omega$
    that contain infinitely many occurences of both $a$ and $b$. Each
    accepting run of the left automaton has to contain infinitely many
    occurrences of both transitions looping on the initial
    state. Hence, there does not exist any automaton with the same
    accepting runs as the left automaton and less than two acceptance
    marks. The right automaton accepts the same language using one
    acceptance mark and a different set of accepting runs.}
  \label{fig:counter}
\end{figure}

@@

Now we describe the construction of the QBF in detail. The
construction is parametrized by two integers $C,K\ge 0$, where $K$ is
the desired number of acceptance marks and $C$ is the number of cubes
of $\psi$. Without loss of generality, we assume that the reduced
automaton will use the acceptance marks $N_K=\{1,2,\ldots,K\}$.
%Further, we consider only subsets $T\subseteq\bigcup_{\textrm{SCC}~S}\delta_S$
We start with a description of Boolean variables used in the
constructed QBF.
\begin{itemize}
\item For each transition $t\in\delta$, variable $e_t$ says whether
  $t$ is in the current set $T$ or not.
  \[e_t=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if }t\in T\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
%  In fact, we consider variables 
\item For each transition $t\in\delta$ and acceptance mark $k\in N_K$,
  variable $n_{t,k}$ says whether $k$ is on the transition
  $\rem(t)$ or not.
  \[n_{t,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if }k\in\mks(\rem(t))\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
\item For each $c\in\{1,2,\ldots,C\}$ and acceptance mark $k\in N_K$,
  variables $i_{c,k}$ and $f_{c,k}$ say whether the $c^\textrm{th}$
  cube of $\psi$ contains atoms $\Inf k$ or $\Fin k$, respectively.
  \[i_{c,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if the $c^\textrm{th}$ cube of $\psi$ contains $\Inf k$}\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
  \[f_{c,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if the $c^\textrm{th}$ cube of $\psi$ contains $\Fin k$}\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]

@@

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G} to get generic TELA) from the Spot
library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA by
\texttt{ltl3tela} with option \texttt{-D1} and by two state-of-the-art
translators from the Owl library~\cite{kretinsky.18.atva}, namely
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} and
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav}. Some
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata set and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. One can observe that a higher level sometimes achieves a smaller
reduction than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl3tela} on automata coming from formulas from
literature). This is caused by the QBF solver timeouts occuring
earlier as formulas constructed by the higher level are more
complex. The automata sets produced by \texttt{ltl2dela} and
\texttt{ltl2dgra}, on formulas from literature do not contain any
automaton where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

@@

\begin{figure}[p]
  \centering
  \input{quantile.tex}
  \bigskip\bigskip
  \input{quantile-random.tex}
  \bigskip
  \caption{Performance of \telatko on individual automata of each automata set.
    Automata sets constructed from formulas from literature are in the upper graph,
    automata sets constructed from random formulas are in the lower graph.
    Each line shows the time ($y$ axis) needed by \telatko to process the
    $x$-th automaton of the set, where automata in the set are ordered by
    their processing time.
    % The $x$ axis represents the number of automata from each automata
    % set that are processed by \telatko in the time shown on the $y$
    % axis.
  }
\label{fig:quantile}\end{figure}

>>>>

  \caption{The left automaton accepts the words % $w\in\{a,b\}^\omega$
    that contain infinitely many occurrences of both $a$ and $b$. Each
    accepting run of the left automaton has to contain infinitely many
    occurrences of both transitions looping on the initial
    state. Hence, there does not exist any automaton with the same
    accepting runs as the left automaton and less than two acceptance
    marks. The right automaton accepts the same language using one
    acceptance mark and a different set of accepting runs.}
  \label{fig:counter}
\end{figure}

@@

Now we describe the construction of the QBF in detail. The
construction is parameterized by two integers $C,K\ge 0$, where $K$ is
the desired number of acceptance marks and $C$ is the number of cubes
of $\psi$. Without loss of generality, we assume that the reduced
automaton will use the acceptance marks $N_K=\{1,2,\ldots,K\}$.
%Further, we consider only subsets $T\subseteq\bigcup_{\textrm{SCC}~S}\delta_S$
We start with a description of Boolean variables used in the
constructed QBF.
\begin{itemize}
\item For each transition $t\in\delta$, variable $e_t$ says whether
  $t$ is in the current set $T$ or not.
  \[e_t=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if }t\in T\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
%  In fact, we consider variables 
\item For each transition $t\in\delta$ and acceptance mark $k\in N_K$,
  variable $n_{t,k}$ says whether $k$ is on the transition
  $\rem(t)$ or not.
  \[n_{t,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if }k\in\mks(\rem(t))\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
\item For each $c\in\{1,2,\ldots,C\}$ and acceptance mark $k\in N_K$,
  variables $i_{c,k}$ and $f_{c,k}$ say whether the $c^\textrm{th}$
  cube of $\psi$ contains atoms $\Inf k$ or $\Fin k$, respectively.
  \[i_{c,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if the $c^\textrm{th}$ cube of $\psi$ contains $\Inf k$}\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]
  \[f_{c,k}=\left\{
      \begin{array}{ll}
        1~~ & \textrm{if the $c^\textrm{th}$ cube of $\psi$ contains $\Fin k$}\\[.5ex]
        0 & \textrm{otherwise}    
      \end{array}\right.
  \]

@@

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parameterized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G} to get generic TELA) from the Spot
library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA by
\texttt{ltl3tela} with option \texttt{-D1} and by two state-of-the-art
translators from the Owl library~\cite{kretinsky.18.atva}, namely
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} and
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav}. Some
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

For each automata set identified by the translator and the set of
formulas, Table~\ref{tab:cumm-data} shows the cumulative numbers of
marks in the input automata set and after each reduction, together
with the reduction ratio and total time spent by the considered
reduction. The column \textit{solver timeout} shows the number of
automata for which the last query to QBF solver did not finish within
the 30~seconds limit. The timeout of the last QBF query means that the
automaton may be potentially further reduced if a longer time limit is
used. One can observe that a higher level sometimes achieves a smaller
reduction than a lower level (e.g., compare level~1 and level~2 for
\texttt{ltl3tela} on automata coming from formulas from
literature). This is caused by the QBF solver timeouts occurring
earlier as formulas constructed by the higher level are more
complex. The automata sets produced by \texttt{ltl2dela} and
\texttt{ltl2dgra}, on formulas from literature do not contain any
automaton where level~2 or level~3 achieves a better result than
level~1. However, all levels contribute to the reductions in the
multi-level setting.

@@

\begin{figure}[p]
  \centering
  \input{quantile.tex}
  \bigskip\bigskip
  \input{quantile-random.tex}
  \bigskip
  \caption{Performance of \telatko on individual automata of each automata set.
    Automata sets constructed from formulas from literature are in the upper graph,
    automata sets constructed from random formulas are in the lower graph.
    Each line shows the time ($y$ axis) needed by \telatko to process the
    $x$-Th automaton of the set, where automata in the set are ordered by
    their processing time.
    % The $x$ axis represents the number of automata from each automata
    % set that are processed by \telatko in the time shown on the $y$
    % axis.
  }
\label{fig:quantile}\end{figure}
